---
title: Sensitivity analysis on the SIR model
author: Stefan Siegert
---

# The SIR model of infectious disease spread

```{r}
sir = function(t, y, parms) {
  with(as.list(c(y, parms)), {
    list(c(-lambda*s*i,
            lambda*s*i - gamma*i,
            gamma*i))
  })
}
```

# Running the model

```{r}
i0 = 1e-6 # initial proportion of population infected
gamma = 1/7 # assume mean recovery time of 7 days
lambda = 2*gamma # assume lambda > gamma
out = deSolve::ode(y = c(s=1-i0, i=i0, r=0),
                   times = seq(0,200,1),
                   func=sir,
                   parms=c(lambda=lambda, gamma=gamma))
out = out %>% `class<-`('matrix') %>% as_tibble
head(out)
```

# Output

```{r}
out %>% gather('compartment', 'fraction', -time) %>%
ggplot(aes(x=time, group=compartment)) + geom_line(aes(y=fraction, colour=compartment))
```

# Import the mogp python module

```{r}
library(reticulate)
use_python('/usr/bin/python3.8')
mogp_emulator = import('mogp_emulator')
```

# Latin hypercube design

```{r}
py_set_seed(42)
n_design = 30
lhc = mogp_emulator$LatinHypercubeDesign(2)
x = lhc$sample(n_design)
```

- scale design points to desired range
- we vary `gamma` in $[1/10, 1/3]$ and `lambda` in $[1/10, 1/3]$


```{r}
x[,1] = x[,1] * (1/3 - 1/10) + 1/10
x[,2] = x[,2] * (1/3 - 1/10) + 1/10
```



# Choosing the 1d output: Number of people affected after 200 days

We use the fraction of previously infected people (i.e. currently infected plus
recovered) after 200 days as the output of interest. 

For now we will use a fixed initial state of $i(0) = 10^{-6}$.

```{r}
f = function(x) {
  i0 = 1e-6
  out = deSolve::ode(y = c(s=1-i0, i=i0, r=0),
                     times = seq(0,200,1),
                     func=sir,
                     parms=c(lambda=x[1], gamma=x[2]))
  return(unname(1 - out[nrow(out), 's']))
}
```


# Run SIR model at these design points and calculate the simulator output

```{r}
fx = apply(x, 1, function(xi) f(xi))
design = cbind(x, fx)
colnames(design) = c('lambda', 'gamma', 'output')
knitr::kable(design)
```

# Building the Gaussian process emulator

We emulate the simulator using a Gaussian process with a linear mean function
and a squared exponential (Gaussian) covariance function.

We fit the emulator using the first 20 design points, and validate using the
remaining 10. 

```{r}
i_train = 1:20
i_validate = 21:30
xx = design[i_train, c('lambda', 'gamma')]
yy = design[i_train, 'output']
gp = mogp_emulator$GaussianProcess(xx, yy)
```


